## Introduction to dplyr

dplyr is an R package in the tidyverse. We can load the package using 

```{r}
install.packages("dplyr")

library(dplyr)
library(ggplot2)
?dplyr

#preview data
head(starwars) #can also do head(starwars,10) for 10 rows instead of standard 6
```

##Filtering data
We can filter data using `filter()`. This allows us to subset observations (rows) based on their values (in columns)

Tips:
  * Be sure you spell the column name correctly (and the value name of it's a categorical variable).     Remember, R is case-sensitive
  * Be sure to use `==` when comparing observations. (Remember, `=` is on assignment operator)
  * You can use >,<,>=,<= to compare numeric or categorical variables (nominal variables are ranked     alphabetically, while ordinal variables have a built-in rank)
```{r}
# or is |   
# and is & or ,
filter(starwars, hair_color=="blond" | hair_color =="blonde")
filter(starwars, hair_color=="blond" & species =="Human")
filter(starwars, species=="Human" , homeworld =="Tatooine", skin_color=="fair") #can use , or &
filter(starwars, hair_color=="blond" , mass >= 75)

filter(starwars, mass!= 75, name<"Mace") #automatically does not include NA values

filter(starwars, mass!=75, is.na(mass), name<"Mace") #includes NA data

filter(starwars, hair_color %in% c("blond", "blonde")) # works very similar to an or sign

importantPlanets <- c("Tatooine", "Naboo", "Hoth")
filter(starwars, homeworld %in% importantPlanets)


#Saving filtered data --- vectors!
filteredData <- filter(starwars, hair_color=="blond" | hair_color =="blonde")

#ranked data
head(diamonds) #displays first 6 rows
class(diamonds$cut) #gives you the specific type of data
summary(diamonds$cut) #gives a count of each category or summary statistics if numeric

summary(diamonds$carat)
head(diamonds$cut) #displays first 6 values (and levels)

filter(diamonds, cut>"Good")

#Ordering categorical data
unique(starwars$eye_color) #returns all the eye colors
factor(starwars$eye_color, c("red", "orange", "gold", "yellow", "green, yellow", "blue", "black"), 
       ordered = T) #ranks the data but does not save it into column if dataset

filter(spotify_data, genre %in% c("rock", "pop", "soul", "country"))

View(starwars)
# Find all characters that are shorter than 100 cm and not a droid
filter(starwars, height<100, species!="Droid")

# Which characters were born between the years 100 and 200 (inclusive)
filter(starwars, birth_year>=100, birth_year<=200)
                #         OR
filter(starwars, between(birth_year, 100, 200))



# Which characters weigh over 100kg but are shorter than 185cm?
filter(starwars, mass>100, height<185)


#Missing value
NA>5 # it doesnt know
NA + 10
NA == NA #returns NA cause it does not know what each NA is...
mass<- NA
mass ==10 #returns NA

is.na(NA) #return true


#How many characters are missing a hair color
filter(starwars, is.na(hair_color))


```


####Arranging data####
`arrange()` reorders rows. It does not remove any rows. NA values are always at the end when you order by a column, no matter which way you arrange it. 
```{r}
#lowest to highest
arrange(starwars, birth_year)

#highest to lowest
arrange(starwars, desc(birth_year))

#categorical is alphabetically
arrange(starwars, hair_color)

# multiple columns
arrange(starwars, hair_color, birth_year)

###practice!

## Arrange starwars characters to find the tallest character and the shortest character
arrange(starwars, desc(height))
arrange(starwars, height)

##Alphabetize the starwars characters by name
arrange(starwars, name)

##How could you use arrange() to sort all missing values to the start?
arrange(starwars, desc(is.na(height)), desc(is.na(mass))...)#keep adding desc(is.na(x)) x=columns for all of x

arrange(starwars, desc(is.na(starwars)))
```

### We can select certain columns in the dataset

`select()` allows us to retain only certain variables (columns). It doesn't change the order, but it removes columns not names 
```{r}

select(starwars, hair_color, skin_color, eye_color)
head(starwars)
select(starwars, hair_color:eye_color) #first:last

select(starwars, -(hair_color:eye_color)) #the -() removes everything inside the () and displays everything elese

starwars_noColor <-select(starwars, -(hair_color:eye_color))
ggplot(starwars_noColor, aes(x=height)) + geom_histogram()


select(starwars, contains("color")) #pulls every column that contains word color
select (starwars, ends_with("color"))#pulls every column that ends with the word color
select (starwars, starts_with("s"), ends_with("color"))


starwars2<-rename(starwars, birthYear = birth_year) #new name = old name

##Practice!

#Select out the homeworld and species for the starwars dataset. What question might this subset of variables help answer?
select(starwars, homeworld, species)

#Brainstorm as many ways as possible to select hair_color, eye_color, skin_color, and birth_year
select(starwars, contains("color"), contains ("year"))
select(starwars, hair_color:birth_year)
select(starwars, ends_with("color"), ends_with("year"))

#We can use select() to reorder columns
select(starwars, species, everything()) #everything() is jus everything else
select(starwars, species, mass, height, name, everything())
```

####Adding new columns####
`mutate()` adds new columns to the end of your dataset.

```{r}

starwars_small <- select(starwars, height, mass, birth_year)

starwars_small <- mutate(starwars_small, height_m = height/100)
mutate(starwars_small, bmi = mass/(height_m^2))

starwars_small <- select(starwars, height, mass, birth_year)
mutate(starwars_small, height_m = height/100, bmi = mass/(height_m^2))#mutate both in one line of code

#to only keep new columns, use transmute()
transmute(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

## using aggregate functions 
prop_mass <- mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = T))
arrange(prop_mass, desc(proportional_mass))
```

####Summarizing and grouping data

`summarize()` collapses an entire column of data to a single value
```{r}

mutate(starwars, mean_mass = mean(mass, na.rm = T))
summarize(starwars, mean_mass = mean(mass, na.rm  = T))

mean(starwars$mass, na.rm = T)

species_masses <- summarize(group_by(starwars, species), 
                    mean_mass = mean(mass, na.rm = T),
                    count = n()) #n(returns the number of values in each species)

arrange(species_masses, desc(mean_mass))            

# the pipe operator %>%
# function (x,y) is the same as %>% function(y)
#When using dplyr functions, generally always start with dataset

species_masses <- starwars %>%
  group_by(species)%>%
  summarize(mean_mass = mean(mass, na.rm=T),
            count = n()) %>%
  arrange(desc(mean_mass))
species_masses

arrange(species_masses, desc(mean_mass))
```


####Sampling a designated #of rows
`sample_n()` allows us to sample a random number of rows from our dataset. Sampling is random
```{r}

#10 random rows
starwars_10rows <- sample_n(starwars, 10)

#10% of rows selected randomly
nrow(starwars) #total number of rows is 87
starwars_10percent <- sample_frac(starwars, 0.1)
starwars_10percent

starwars %>%
  group_by(gender) %>%
  sample_n(4) #4 random character per gender 
```

We can also take a 'slice' of our dataset using `slice()` and its related set of functions. This retains the order of our dataset and is not random.
```{r}

slice(starwars, 50:60) #slice by index value (row number)

slice_head(starwars, n=10) #first 10 rows by index
slice_tail(starwars, n=10) #last 10 rows by index

slice_min(starwars, order_by = height, n=10) #rows with lowest 10 masses
slice_max(starwars, order_by = mass, n=10) #rows with highest 10 masses

```

